---
title: "Notes on Chapter 6"
author: "The Caveman Coder"
date: "2025-12-31"
output: html_document
---

# Chapter 6: Spatial Modeling of Areal Data. Lip Cancer in Scotland.

Aim: Estimate the risk of lip cancer in Scortland, UK, and the proportion of the population engaged in agriculture, fishing, or forestry (AFF) for each of the Scottish counties, using the **R-INLA** package.  Data are obtained from the **SpatialEpi** package.  


## Data and map

Loading the **SpatialEpi** package and the relevant data for Scotland:
```{r}
library(SpatialEpi)
data(scotland)

class(scotland)
```

Inspecting the Scotland data:
```{r}
names(scotland)
```

Data dictionary for `scotland`:
- `geo`: datafreame with names and centroid coordinates (eastings/northings) of each of the counties  
- `data`: dataframe with names, number of observed and expected lip cancer cases, and AFF values of each of the countries  
- `spatial.polygon`: `SpatialPolygons` object with the map of Scotland  
- `polygon`: polygon map of Scotland  

```{r}
head(scotland$data)
```

Plotting the map of Scotland:
```{r}
map <- scotland$spatial.polygon
plot(map)
```

This `map` does not contain information about the CRS. 

Specifying a CRS for `map` manually:
```{r}
proj4string(map) <- "+proj=tmerc +lat_0=49 +lon_0=-2 +k=0.9996012717 +x_0=400000 +y_0=-100000 +datum=OSGB36 +units=km +no_defs"
```

For more details on how to do this, see https://spatialreference.org/ref/epsg/27700/.

Preparing the map data for mapping using the **leaflet** package:
```{r}
library(sf)
map <- st_as_sf(map)
map <- st_transform(map, crs = 4326)
```

The snippet above is necessary because **leaflet** expects the data to be specified in latitude and longitude using WGS84. The map was first transformed from `sp` to `sf` object and then `st_transform()` was used to convert the coordinates.  


## Data preparation

Creating a dummy dataframe called `d` with columns containing the county ids, the observed and expected number of lip cancer cases, the AFF values, and the SIRs:
```{r}
d <- scotland$data[, c("county.names", "cases", "expected", "AFF")]
names(d) <- c("county", "Y", "E", "AFF")

head(d)
```

Calculating the SIR using the formula from Chapter 5:
```{r}
d$SIR <- d$Y / d$E
head(d)
```

### Adding data to map

Joining `d` with `map` using `cbind`:
```{r}
map <- cbind(map, d)
head(map)
```


## Mapping SIRs

Visualizing the observed and expected lip cancer cases using an interactive choropleth map in **leaflet**:
```{r fig.cap="Interactive map of lip cancer SIR in Scotland counties created with **leaflet**."}
library(leaflet)

l <- leaflet(map) |> 
  addTiles()

pal <- colorNumeric(palette = "YlOrRd", domain = map$SIR)

l |> addPolygons(
  color = "grey", weight = 1,
  fillColor = ~ pal(SIR), fillOpacity = 0.5
) |> 
  addLegend(
    pal = pal, values = ~ SIR, opacity = 0.5,
    title = "SIR",
    position = "bottomright"
  )
```

Adding "mouse hover effects" to the interactive map:
```{r}
labels <- sprintf(
  "<strong> %s </strong> <br/>
  Observed: %s <br/> Expected: %s <br/>
  AFF: %s <br/> SIR: %s",
  map$county, map$Y, round(map$E, 2), map$AFF, round(map$SIR, 2)
) |> 
  lapply(htmltools::HTML)

l |> addPolygons(
  color = "grey", weight = 1,
  fillColor = ~ pal(SIR), fillOpacity = 0.5,
  highlightOptions = highlightOptions(weight = 4),
  label = labels,
  labelOptions = labelOptions(
    style = list(
      "font-weight" = "normal",
      padding = "3px 8px"
    ),
    textsize = "15px",
    direction = "auto"
  )
) |> addLegend(
  pal = pal, 
  values = ~ SIR,
  opacity = 0.5,
  title = "SIR",
  position = "bottomright"
)
```

The "mouse-hover" effects are created by adding the arguments `highlightOptions`, `label`, and `labelOptions` to `addPolygons()`. The labels are created using HTML syntax inside the `sprintf` function, which returns a character vector containing  a formatted combination of text and variables. The function `htmltools::HTML()` marks the text as HTML. 


## Modeling

We specify a model assuming that the observed counts, $Y_i$, are conditionally independently Poisson distributed:
$$
Y_i \sim Poisson(E_i \theta_i), \space i = 1, \space ..., \space n
$$

where $E_i$ is the expected count and $\theta_i$ is the relative risk in area $i$...

### Neighborhood matrix

Creating the neighborhood matrix need to define the spatial random effect using `poly2nb()` and `nb2INLA()` functions from the **spdep** package:
```{r}
library(spdep)
library(INLA)

nb <- poly2nb(map)
head(nb)
```

Using `nb2INLA()` to convert the `nb` list into a file with representation of the neighborhood matrix as required by the **R-INLA** package. 
```{r}
nb2INLA("map.adj", nb)
```

Reading the converted list using `inla.read.graph()`:
```{r}
g <- inla.read.graph(filename = "map.adj")
```

### Inference using INLA

Since the model includes two random effects, $u_i$ (for modeling the residual variation) and $v_i$ (for modeling unstructured noise), we need to include two vectors in the data that denote the indices of these random effects.

Creating the `idareau` and `idareav` variables that hold the indices for the random effects:
```{r}
map$idareau <- 1:nrow(map)
map$idareav <- 1:nrow(map)
```

Specifying the model formula:
```{r}
formula <- Y ~ AFF + 
  f(idareau, model = "besag", graph = g, scale.model = TRUE) +
  f(idareav, model = "iid")
```

In the snipped avbove, the response variable is $Y$ and the covariate is $AFF$. The random effects are defined using `f()`. For $u_i$, the model `"besag"` was used, with the neighborhood matrix given by `g`. The option `scale.model = TRUE` makes the precision parameter of models with different CAR priors comparable. For $v_i$, `model = "iid"` was chosen.  

Fitting the model by calling the `inla()` function:
```{r}
res <- inla(
  formula,
  family = "poisson",
  data = map,
  E = E,
  control.predictor = list(compute = TRUE),
  control.compute = list(return.marginals.predictor = TRUE)
)
```

### Results

Inspecting the results object using `summary()`:
```{r}
summary(res)
```

The results indicate that AFF increases lip cancer risk. 

Plotting the posterior distribution of the coefficient using `inla.smarginal()` and `ggplot()`:
```{r fig.cap="Posterior distribution of the coefficient of covariate AFF."}
library(ggplot2)

marginal <- inla.smarginal(res$marginals.fixed$AFF)
marginal <- data.frame(marginal)

ggplot(marginal, aes(x = x, y = y)) +
  geom_line() +
  labs(
    x = expression(beta[1]),
    y = "Density"
  ) +
  geom_vline(xintercept = 0, col = "black") +
  theme_bw()
```


## Mapping relative risks

Inspecting the estimates of the relative risk of lip cancer and their uncertainty for each of the counties:
```{r}
head(res$summary.fitted.values)
```

The column `mean` is the mean posterior and the ` 0.025quant` and `0.975quant` are the 2.5th and 97.5th quantiles, respectively. 

Adding these data to `map` to make plotting easier (and the code more readable):
```{r}
map$RR <- res$summary.fitted.values[, "mean"]
map$LL <- res$summary.fitted.values[, "0.025quant"]
map$UL <- res$summary.fitted.values[, "0.975quant"]
```

Showing the relative risk of lip cancer in the interactive leaflet map:
```{r fig.cap="Interactive map of lip cancer relative risk in Scotland counties created with **leaflet**."}
pal <- colorNumeric(palette = "YlOrRd", domain = map$RR)

labels <- sprintf(
  "<strong> %s </strong> <br/>
  Observed: %s <br/> Expected: %s <br/>
  AFF: %s <br/> SIR: <br/> RR: %s (%s, %s)",
  map$county, map$Y, round(map$E, 2), round(map$AFF, 2), round(map$RR, 2), round(map$LL, 2), round(map$UL, 2)
) |> lapply(htmltools::HTML)

lRR <- leaflet(map) |> 
  addTiles() |> 
  addPolygons(
    color = "grey",
    weight = 1,
    fillColor = ~ pal(RR),
    fillOpacity = 0.5,
    highlightOptions = highlightOptions(weight = 4),
    label = labels,
    labelOptions = labelOptions(
      style = list(
        "font-weight" = "normal",
        padding = "3px 8px"
      ),
      textsize = "15px", direction = "auto"
    )
  ) |> 
  addLegend(
    pal = pal,
    values = ~ RR,
    opacity = 0.5,
    title = "RR"
  )

lRR
```


## Exceedance probabilities

Probabilities of relative risk estimates being greater than a given threshold are called exceedance probabilities. These are useful to assess unusual elevation of disease risk.  Mathematically, 
$$
P(\theta_i > c) = 1 = P(\theta_i \leq c)
$$
in **R-INLA**, calculation is as follows:
```r
1 - inla.pmarginal(q = c, marginal = marg)
```
where `marg` is the marginal distribution of the predictions, and `c` is the threshold value. 

In our example, the marginals of the relative risks azre in `res$marginals.fitted.values`, and the marginal corresponding to the first county is in `res$marginals.fitted.values[[1]]`. 

Calculating the probability that the relative risk of the first county exceeds 2 (or $P(\theta_1 > 2)$):
```{r}
marg <- res$marginals.fitted.values[[1]]
exc_prob_2 <- 1 - inla.pmarginal(q = 2, marginal = marg)
exc_prob_2
```

calculating the exceedance probabilities for all counties using the `sapply()` function:
```{r}
exc <- sapply(
  res$marginals.fitted.values,
  FUN = function(marg){1 - inla.pmarginal(q = 2, marginal = marg)}
)

head(exc)
```

Adding the exceedance probabilities to `map`:
```{r}
map$exc <- exc
```

Creating an interactive map of the exceedance probabilities using **leaflet**:
```{r fig.cap="Interactive map of exceedance probabilities in Scotland counties created with **leaflet**."}
pal <- colorNumeric(palette = "YlOrRd", domain = map$exc)

labels <- sprintf(
  "<strong> %s </strong> <br/>
  Observed: %s <br/> Expected: %s <br/>
  AFF: %s <br/> SIR: %s <br/> RR: %s (%s %s) <br/> P(RR>2): %s",
  map$county, map$Y, round(map$E, 2), map$AFF,
  round(map$SIR, 2), round(map$RR, 2), round(map$LL, 2),
  round(map$UL, 2), round(map$exc, 2)
) |> lapply(htmltools::HTML)

lexc <- leaflet(map) |> 
  addTiles() |> 
  addPolygons(
    color = "grey",
    weight = 1,
    fillColor = ~ pal(exc),
    fillOpacity = 0.5,
    highlightOptions = highlightOptions(weight = 4),
    label = labels,
    labelOptions = labelOptions(
      style = list(
        "font-weight" = "normal",
        padding = "3px 8px"
      ),
      textsize = "15px",
      direction = "auto"
    )
  ) |> 
  addLegend(
    pal = pal,
    values = ~ exc,
    opacity = 0.5,
    title = "P(RR>2)",
    position = "bottomright"
  )

lexc
```






