---
title: "Notes on Chapter 5"
author: "The Caveman Coder"
date: "2025-12-30"
output: html_document
---

# Areal Data

Areal or lattice data arise when a fixed domain is partitoned into a finite number of subregions at which outcomes are aggregated.  

Examples:
- The number of cancer cases in countries  
- The number of road accidents in provinces  
- The proportion of people living in poverty in census tracts  


## Spatial neighborhood matrices

Example showing how neighbhors of several counties in the map of Pennsylvania, USA are coumputed using a neighborhood definition based on common boundaries of the counties. 

Showing first the map of Pennsylvania obtained from the **SpatialEpi** package:
```{r fig.cap="Map of Pennsylvania counties."}
library(SpatialEpi)

map <- pennLC$spatial.polygon
plot(map)
```

Check what type of object `map` is:
```{r}
class(map)
```


Obtaining the neighbors of each county in the map using the `poly2nb()` function from the **spdep** package:
```{r}
library(spdep)

nb <- poly2nb(map)
head(nb)
```

The `poly2nb` function returns a neighbors list based on counties with contiguous boundaries. Each element in `nb` represents one county and contains the indices of its neighbors.  

Showing the neighbors of specific counties of Pennsylvania, say neighbors of counties 2, 44, and 58:
```{r}
d <- data.frame(county = names(map), neigh = rep(0, length(map)))

head(d)
rownames(d) <- names(map)
```

Continuing... Creating a `SpatialPoligonsDataFrame` object from the original `map` and `d`:
```{r}
map <- SpatialPolygonsDataFrame(map, d, match.ID = TRUE)
```

Continuing... Setting `neigh` equal to 1 for counties that are neighbors of counties 2, 44, and 58 (otherwise, 0):
```{r}
map$neigh[nb[[2]]] <- 1
map$neigh[nb[[44]]] <- 1
map$neigh[nb[[58]]] <- 1
```

Adding coordinates to our data, plus a variable called `ID` that contains the `id` of the counties:
```{r}
coord <- coordinates(map)
map$long <- coord[, 1]
map$lat <- coord[, 2]
map$ID <- 1:dim(map@data)[1]
```

Plotting the map with `ggplot()`. First, convert `map` (which is a `SpatialPolygonsDataFrame` object) to a simple feature object of class `sf` using `st_as_sf()` function from the **sf** package:
```{r fig.cap="Neighbors of areas 2, 44, and 58 of Pennsylvania."}
library(sf)
library(ggplot2)

mapsf <- st_as_sf(map)

ggplot(mapsf) +
  geom_sf(aes(fill = as.factor(neigh))) +
  geom_text(aes(x = long, y = lat, label = ID), color = "white") +
  theme_bw() +
  guides(fill = "none")
```


## Standardized incidence ratio

This is the ratio of observed counts to the expected counts:
$$
SIR_i = Y_i / E_i
$$
The expected counts $E_i$ represent the total number of cases that one would expect if the population if area *i* behaved in a way the standard (or regional) population behaves.  

$E_i$ can be calculated using:
$$
E_i = \sum_{j=1}^{m} r_j^{(s)} n_j^{(i)}
$$
where $r_j^{(s)}$ is the rate or number of cases divided by the population in $j$ in the standard population, and $n_j^{(i)}$ is the population stratum $j$ of area $i$.

The number of lung cancer cases and the population of Pennsylvania at country level stratified on race (white or non-white), gender (female or male), and age (under 40, 40-59, 60-69, and 70+) can be found in `pennLC$data`. 
```{r}
head(pennLC$data)
```

Continuing... 
```{r}
library(dplyr)

d <- pennLC$data |> 
  group_by(county) |> 
  summarize(Y = sum(cases))

head(d)
```

Calculating the expected number of cases in each county using indirect standardization using the `expected()` function of **SpatialEpi**:
```{r}
# First sort the data
pennLC$data <- pennLC$data[order(
  pennLC$data$county,
  pennLC$data$race,
  pennLC$data$gender,
  pennLC$data$age
), ]

# Using the `expected()` function
E <- expected(
  population = pennLC$data$population,
  cases = pennLC$data$cases,
  n.strata = 16
)

head(E)
```

Adding vector `E` to our data frame `d` where the `E` elements correspond to the counties in `d$country` in the same order:
```{r}
d$E <- E[match(d$county, unique(pennLC$data$county))]
head(d)
```

Note the the function `match()` was used to calculate the vector of the positions that match `d$county` in `unique(pennLC$data$county)`, which are the corresponding counties of `E`.

Computing the SIR values (ratio of the observed to the expected counts) using the formula explained above:
```{r}
d$SIR <- d$Y / d$E
head(d)
```

Adding the SIR info back to our `map` object:
```{r}
map <- merge(map, d)
```

Converting the `map` object (a `SpatialPolygonDataFrame` object) to an `sf` object, in preparation for plotting:
```{r}
mapsf <- st_as_sf(map)
```

Plotting the map using `ggplot()` (now with SIR info used to color the counties):
```{r fig.cap="SIR of lung cancer in Pennsylvania counties."}
mapsf |> ggplot() +
  geom_sf(aes(fill = SIR)) +
  scale_fill_gradient2(
    midpoint = 1, 
    low = "blue",
    mid = "white",
    high = "red"
  ) +
  theme_bw()
```

The white color fill represents counties where the number of lung cancer cases observed match the number of expected cases. In counties where SIR > 1 or the number of lung cancer cases observed is higher than the number of expected cases, the fill color is reddish. In counties where SIR < 1 or the number of observed lung cancer cases is lower than that of expected cases, the fill color is bluish.  


## Spatial small area disease risk estimation

In regions with small populations or rare diseases are expected, the expected counts may be very low, and this leads to misleading SIRs. 

A popular spatial model that we can use for this case is the BYM model (Besag-York-Mollie). In this model, the spatial random effect is assigned a Conditional Autoregressive distribution which wmoothes the data according to certain neighborhood structure that specifies that two areas are neighbors if they share a common boundary.  

In **R-INLA** the BYM model can be accessed by specifying the formula in this manner:
```r
formula <- Y ~ 
  f(idareeau, model = "besag", graph = g, scale.model = TRUE) +
  f(idareav, model = "iid")
```

In this formula, the option `scale.model = TRUE` is used to make the precision parameter of models with different CAR priors comparable.  

A new parametrization of the BYM model called BYM2 was proposed by Simpson et al. in 2017. In **R-INLA**, the BMY2 model can be specified by creating the formula like this:
```r
formula <- Y ~ f(idarea, model = "bym2", graph = g)
```

### Spatial modeling of lung cancer in Pennsylvania

Using the BYM2 model to calculate relative risks of lung cancer in the Pennsylvania counties:

First we define the `idarea` variable (typically a vector `c(1, 2, ..., I)` where `I` is the number of counties).
```{r}
map$idarea <- 1:nrow(map@data)
```

Then we define a PC prior for the marginal precision...
```{r}
prior <- list(
  prec = list(
    prior = "pc.prec",
    param = c(0.5 / 0.31, 0.01)
  ),
  phi = list(
    prior = "pc",
    param = c(0.5, 2 / 3)
  )
)
```

Computing the object `g` with the neighborhood matrix that will be used in the spatially structured effect (create first a neighbors list `nb` using `poly2nb()` then use `nb2INLA()` to convert it to a neighborhood representation matrix required by the **R-INLA** package):
```{r}
# creation of neighbors list nb
nb <- poly2nb(map)
head(nb)

# creation of neighborhood representation matrix
library(INLA)

nb2INLA("map.adj", nb)
g <- inla.read.graph(filename = "map.adj")
```

Specifying the formula:
```{r}
formula <- Y ~ f(idarea, model = "bym2", graph = g, hyper = prior)
```

Fitting the model by calling the `inla()` function:
```{r}
res <- inla(
  formula, 
  family = "poisson",
  data = map@data,
  E = E,
  control.predictor = list(compute = TRUE)
)
```

The parameter `control.predictor = TRUE` means that we set the `control.predictor` parameter equal to `list(compute = TRUE`) to compute the posteriors of the predictions.

Checking the results:
```{r}
summary(res)
```

Looking at the summaries of relative risks:
```{r}
head(res$summary.fitted.values)
```

Adding columns to the results in preparation for plotting:
```{r}
map$RR <- res$summary.fitted.values[, "mean"]
map$LL <- res$summary.fitted.values[, "0.025quant"]
map$UL <- res$summary.fitted.values[, "0.975quant"]

summary(map@data[, c("RR", "LL", "UL")])
```

Code for generating plot objects:
```{r}
mapsf <- st_as_sf(map)

gRR <- mapsf |> 
  ggplot() +
  geom_sf(aes(fill = RR)) +
  scale_fill_gradient2(
    midpoint = 1,
    low = "blue", 
    mid = "white",
    high = "red",
    limits = c(0.7, 1.5)
  ) +
  theme_bw()

gRR
```

for gLL:
```{r}
gLL <- mapsf |> 
  ggplot() +
  geom_sf(aes(fill = LL)) +
  scale_fill_gradient2(
    midpoint = 1,
    low = "blue", 
    mid = "white",
    high = "red",
    limits = c(0.7, 1.5)
  ) + 
  theme_bw()

gLL
```

for gUL:
```{r}
gUL <- mapsf |> 
  ggplot() +
  geom_sf(aes(fill = UL)) +
  scale_fill_gradient2(
    midpoint = 1,
    low = "blue", 
    mid = "white",
    high = "red",
    limits = c(0.7, 1.5)
  ) + 
  theme_bw()

gUL
```

Plotting the three objects side-by-side using the `plot_grid()` function from the **cowplot** package:
```{r}
library(cowplot)
plot_grid(gRR, gLL, gUL, ncol = 1)
```
 
Of course, we could have used `facet_grid()` or `facet_wrap()` to achieve the same effect. 


Looking at the summary of random effects in the results:
```{r}
head(res$summary.random$idarea)
```

Adding another column to the results to make plotting easier:
```{r}
mapsf$re <- res$summary.random$idarea[1:67, "mean"]
head(mapsf)
```

Visualizing the posterior mean of the random effects:
```{r fig.cap="Posterior mean of the BYM2 random effect."}
mapsf |> ggplot() +
  geom_sf(aes(fill = re)) +
  scale_fill_gradient2(
    midpoint = 0,
    low = "blue",
    mid = "white",
    high = "red"
  ) +
   theme_bw()
```


