---
title: "Notes on Chapter 7"
author: "The Caveman Coder"
date: "2025-12-31"
output: html_document
---

# Chapter 7: Spatio-temporal modeling of areal data.

Aim: Estimate risk of lung cancer in Ohio using the **R-INLA** pakage and visualize data using **ggplot2**, **plotly**, and **gganimate**. Data are obtained from the **SpatialEpiApp** package.  

Note: **SpatialEpiApp** is no longer available in CRAN. Manually installing the package is also problematic due to deprecated dependencies. 

## Data and map

Loading the data source `dataohiocomplete.csv`:
```{r}
dohio <- read.csv("data/dataohiocomplete.csv")
head(dohio)
```

Loading the shafefile:
```{r fig.cap="Map of Ohio counties"}
library(sf)

map <- st_read("data/fe_2007_39_county/fe_2007_39_county.shp")
plot(map[1])
```


## Data preparation

The data contain the number of lung cancer cases and population stratified by gender and race in each of the Ohio counties from 1968 to 1988.


### Observed cases

Obtaining the number of cases for all the strata in each county and year:
```{r}
library(dplyr)

head(dohio)
```

```{r}
d <- aggregate(
  x = dohio$y,
  by = list(county = dohio$NAME, year = dohio$year),
  FUN = sum
)

names(d) <- c("county", "year", "Y")
head(d)
```

Or in **dplyr**:
```{r}
dohio |> 
  select(NAME, year, y) |> 
  group_by(year, NAME) |> 
  summarize(Y = sum(y)) |> 
  rename(county = NAME)
```

### Expected cases

To calculate the expected number of cases using indirected standardization with the `expected()` function of the **SpatialEpi** package. But before doing that, we need to sort the values of the data `dohio` which is required by the `expected()` function. 

Sorting the values of `dohio`:
```{r}
dohio <- dohio[order(
  dohio$county,
  dohio$year,
  dohio$gender,
  dohio$race
), ]
```

Inspecting the first rows ot the sorted data:
```{r}
dohio[1:20, ]
```

Calculating the expected number of cases using the `expected()` function:
```{r}
library(SpatialEpi)

n.strata <- 4
E <- expected(
  population = dohio$n,
  cases = dohio$y,
  n.strata = n.strata
)
```

Now we need to create a new dataframe with the expected counts for each county and year. We will call this `dE`. To do this, we need to detemine the number of years and create a vector of counties where each county repeats `nyears` times:
```{r}
nyears <- length(unique(dohio$year))
countiesE <- rep(unique(dohio$NAME), each = nyears)
```

The corresponding years of `E` are obtained by repeating the whole vector of years `ncounties` times:
```{r}
ncounties <- length(unique(dohio$NAME))
yearsE <- rep(unique(dohio$year), times = ncounties)
```

Finally, the new dataframe `dE` can be created:
```{r}
dE <- data.frame(county = countiesE, year = yearsE, E = E)

head(dE)
```

We need to merge this with `d`.

Merging `dE` with `d`:
```{r}
d <- merge(d, dE, by = c("county", "year"))
head(d)
```

### SIRs

Calculating the SIR or county $i$ and year $j$:
```{r}
d$SIR <- d$Y / d$E
head(d)
```

### Adding data to map

We need to add the data `d` to the `map` object. However, data `d` is in long format. Solution: transform data `d` from long format to wide format, then add to `map`.

Transforming data `d` from long to wide format using the `reshape()` function:
```{r}
dw <- reshape(
  d, 
  timevar = "year",
  idvar = "county",
  direction = "wide"
)

dw[1:2, ]
```

Viewing the original map first:
```{r}
map[1:2, ]
```

Merging the `map` and `dw` using `left_join()`:
```{r}
map <- dplyr::left_join(map, dw, by = c("NAME" = "county"))

map[1:2, ]
```


## Mapping SIRs

To plot `map` object, we need to transform the data from wide to long format. The **tidyr** function `gather()` is useful here:
```{r}
library(tidyr)
map <- gather(map, year, SIR, paste0("SIR.", 1968:1988))
```

Converting the values generated by the `paste0` function above to integers (aka years):
```{r}
map$year <- as.integer(substring(map$year, 5, 8))
```

Plotting with faceting:
```{r fig.cap="Maps of lung cancer SIR in Ohio counties from 1968 to 1988 created with a diverging color scale."}
library(ggplot2)

map |> ggplot() +
  geom_sf(aes(fill = SIR)) +
  facet_wrap(~year, dir = "h", ncol = 7) +
  ggtitle("SIR") +
  theme_bw() +
  theme(
    axis.text.x = element_blank(),
    axis.text.y = element_blank(),
    axis.ticks = element_blank()
  ) +
  scale_fill_gradient2(
    midpoint = 1,
    low = "blue",
    mid = "white",
    high = "red"
  )
```


## Time plots of SIRs

Plotting the SIRs of each of the counties over time:
```{r fig.cap="Time plot of lung cancer SIR in Ohio counties from 1968 to 1988."}
g <- ggplot(
  d, aes(x = year, y = SIR, group = county, color = county)
 ) +
  geom_line() +
  geom_point(size = 2) + 
  theme_bw()

g
```

Removing the legend from plot:
```{r}
g <- g + theme(legend.position = "none")
g
```

Highlighting the time series plot of a specific county (say, Adams) using the `gghighlight()` function from the **gghighlight** package:
```{r}
library(gghighlight)

g + gghighlight(county == "Adams")
```

Transforming the plot to an interactive one using the **plotly** package:
```{r}
library(plotly)

ggplotly(g)
```


## Modeling

Spatio-temporal modeling using **R-INLA** to obtain disease risk estimates.

### Model 

The Bernardinelli model will be used to estimate the relative risk of lung cancer for each Ohio county and year. This model ass;umes that the number of cases $Y_{ij}$ observed in county $i$ and year $j$ are modeled as:
$$
Y_{ij} \sim Po(E_{ij}\theta_{ij}),
$$
where $Y_{ij}$ us the observed number of cases, $E_{ij}$ is the expected number of cases, and $\theta_{ij}$ is the relative risk of county $i$ and year $j$...

### Neighborhood matrix

Creating a neighborhood matrix (necessary to define the spatial random effect) using `poly2nb()` and `nb2INLA()` functions from the **spdep** package:

```{r}
library(INLA)
library(spdep)

nb <- poly2nb(map)
head(nb)
```

Converting the matrix into the format required by INLA:
```{r}
nb2INLA("map.adj", nb)
```

Reading the newly-created graph using `inla.read.graph()`:
```{r}
g <- inla.read.graph(filename = "map.adj")
```


### Inference using INLA

Creating the index vectors for the counties and years that will be used to specify the random effects of the model:
```{r}
d$idarea <- as.numeric(as.factor(d$county))
d$idarea1 <- d$idarea
d$idtime <- 1 + d$year - min(d$year)
```

Note that the vector `idarea` was replicated. Reason: we need to use the index vector of the areas in two different random effects, and **R-INLA** variables can only be associated with an `f()` function only once.  

Formula for the Bernardinalli model:
$$
Y_{ij} \sim Po(E_{ij}\theta_{ij}) \\
\log(\theta_{ij}) = \alpha + u_i + v_i + (\beta + \delta_i) \times t_j
$$
Writing the formula in R-code:
```{r}
formula <- Y ~ f(idarea, model = "bym", graph = g) +
  f(idarea1, idtime, model = "iid") + idtime
```

The first `f()` corresponds to $u_i + v_i$ and the second `f()` is the $\delta_i \times t_j$ part.

Calling the `inla()` function to run the model and generate results:
```{r}
res <- inla(
  formula,
  family = "poisson",
  data = d, 
  E = E,
  control.predictor = list(compute = TRUE)
)
```

Again, the `control.predictor` argument is set to `list(compute = TRUE)` to include the posterior means of the predictors in the results. 


## Mapping relative risks

Extracting the relative risk estimates and lower and upper limits of the 95% credible intervals and adding them to `d` to make mapping easier and the code more readable:
```{r}
d$RR <- res$summary.fitted.values[, "mean"]
d$LL <- res$summary.fitted.values[, "0.025quant"]
d$UL <- res$summary.fitted.values[, "0.975quant"]
```


Merging `d` with `map` to include the columns `RR`, `LL`, and `UL` in the plot:
```{r}
map <- merge(
  map, d,
  by.x = c("NAME", "year"),
  by.y = c("county", "year")
)
```

Using `ggplot` to create the plot:
```{r fig.cap="Maps of lung cancer relative risk in Ohio conties from 1968 to 1988."}
map |> ggplot() +
  geom_sf(aes(fill = RR)) +
  facet_wrap(~ year, dir = "h", ncol = 7) +
  ggtitle("RR") +
  theme_bw() +
  theme(
    axis.text.x = element_blank(),
    axis.text.y = element_blank(),
    axis.ticks = element_blank()
  ) +
  scale_fill_gradient2(
    midpoint = 1,
    low = "blue",
    mid = "white",
    high = "red"
  )
```


Using **gganimate** to create an animated version of this map, instead of faceting by year:
```{r fig.cap="Animation of lung cancer relative risk in Ohio counties from 1968 to 1988."}
library(gganimate)

p <- map |> ggplot() +
  geom_sf(aes(fill = RR)) + 
  theme_bw() +
  theme(
    axis.text.x = element_blank(),
    axis.text.y = element_blank(),
    axis.ticks = element_blank()
  ) + scale_fill_gradient2(
    midpoint = 1,
    low = "blue",
    mid = "white",
    high = "red"
  ) +
  transition_time(year) +
  labs(title = "Year: {round(frame_time, 0)}")
  
p
```

Saving the animation:
```r
anim_save(
  "OhioLungCancerRR.gif", 
  animation = p, 
  width = 800,
  height = 600, 
  res = 100,
  fps = 20
)
```







