---
title: "Notes on Chapter 10"
author: "The Caveman Coder"
date: "2026-01-02"
output: html_document
---


# Chapter 10 - Spatio-temporal modeling of geostatistical data: Air pollution in Spain

Aim: Show how to fit a spatio-temporal model to predict the particulate air pollution levels ($PM_{2.5}$) in Spain from data collected from 2015 to 2017 using the stochastic partial differential equation (SPDE) in the **R-INLA** package.  


## Map

Obtaining the map of Spain using the `getData()` function from the **raster** package:
```{r, fig.cap="Map of Spain."}
library(lwgeom)
library(geodata)

# getData is removed. Using `geodata` 
m <- gadm("Spain", level = 0)


plot(m)
```

Note: `getData()` is deprecated. Used `geodata::gadm()` instead. 

Removing the islands from the map (retaining the Spanish mainland only):
```{r}
library(sf)
library(dplyr)

m <- m |> st_as_sf() |> 
  st_cast("POLYGON")

m$area <- st_area(m) 
m <- m |> arrange(desc(area)) |> 
  slice(1)
head(m)
```

Plotting the Spanish mainland:
```{r, fig.cap="Map of Spain without islands."}
library(ggplot2)

ggplot(m) +
  geom_sf() +
  theme_bw()
```

Transforming the map coordinate system to UTM projection (because we intend to work with meters):
```{r, fig.cap="Map of Spain without islands and with UTM projection."}
m <- m |> st_transform(25830)
m |> ggplot() +
  geom_sf() +
  coord_sf(datum = st_crs(m)) +
  theme_bw()
```

Note: the ESPG code for Spain is 25830.

I cannot find the data for this chapter. (dataPM25.csv)




