---
title: "Notes on Chapter 12"
author: "The Caveman Coder"
date: "2026-01-03"
output: html_document
---

# Building a dashboard to visualize spatial data with **flexdashboard**

Dashboards are tools for effective data visualization that help communicate information in an intuitive and insightful manner,and are essential to support data-driven decision making.  


## The R package **flexdashboard**

Brief review of R Markdown...

### R Markdown

- allows easy work reproducibility by including R code that generates results, and narrative text explaining the work. 

When the Markdown file is complied, the R code is executed and the results are appended to a report that can take a variety of formats including HTML and PDF.

### Layout

xxx

### Dashboard components

A flexdashboard can include these components:
- Interactive JavaScript data visualizations based on HTML widgets  
- Charts created with standard R graphics  
- Simple tables created with `knitr::kable()`  
- Interactive tables created with the **DT** package  
- Value boxes created with the `valueBox()` function  
- Gauges that display values on a meter  
- Text, images, and equations  
- Navigation bar with links to source code, websites, and other tools related to the dashboard.  

## A dashboard to visualize global air pollution

### Data

Map data for this exercise come from **rnaturalearth**.
```{r, fig.cap="World map obtained from the **rnaturalearth** package."}
library(sf)
library(rnaturalearth)

map <- ne_countries(returnclass = "sf")
names(map)[names(map) == "iso_a3"] <- "ISO3"
names(map)[names(map) == "name"] <- "NAME"

head(map)
plot(map)
```

Obtaining $PM_{2.5}$ concentration levels from the **wbstats** package:
```{r}
library(wbstats)
indicators <- wb_search(pattern = "pollution")

head(indicators)
```

Downloading the 2016 data whose `indicator_id` is `EN.ATM.PM25.MC.M3` using the `wb_data()` function:
```{r}
d <- wb_data(
  indicator = "EN.ATM.PM25.MC.M3",
  start_date = 2016,
  end_date = 2016
)

head(d)
```

Assigning the $PM_{2.5}$ readings from `d` to `map` where the country matches, using the `match()` function:
```{r}
map$PM2.5 <- d[match(map$ISO3, d$iso3c), ]$EN.ATM.PM25.MC.M3
head(map)
```


### Table using **DT**

Creating an interactive table using the `datatable()` function from the **DT** package:
```{r, fig.cap="Table with the $PM_{2.5}$ values."}
library(DT)

DT::datatable(
  map[, c("ISO3", "NAME", "PM2.5")],
  rownames = FALSE,
  options = list(pageLength = 10)
)
```


### Map using **leaflet**

Creating an interactive map using **leaflet**:
```{r, fig.cap="Leaflet map with the $PM_{2.5}$ values."}
library(leaflet)

pal <- colorBin(
  palette = "viridis", domain = map$PM2.5,
  bins = seq(0, max(map$PM2.5, na.rm = TRUE) + 10, by = 10)
)


map$labels <- paste0(
  "<strong> Country: </strong> ",
  map$NAME, "<br/> ",
  "<strong> PM2.5: </strong> ",
  map$PM2.5, "<br/> "
) %>%
  lapply(htmltools::HTML)

leaflet(map) |> 
  addTiles() |> 
  setView(lng = 0, lat = 30, zoom = 2) |> 
  addPolygons(
    fillColor = ~ pal(PM2.5),
    color = "white",
    fillOpacity = 0.7,
    label = ~labels,
    highlight = highlightOptions(
      color = "black",
      bringToFront = TRUE
    )
  ) |> 
  leaflet::addLegend(
    pal = pal, values = ~PM2.5,
    opacity = 0.7, title = "PM2.5"
  )
```

### Histogram using **ggplot2**

Creating a histogram with the $PM_{2.5}$ values using `ggplot()`:
```{r, fig.cap="Histogram of the $PM_{2.5}$ values."}
library(ggplot2)

map |> ggplot(
  aes(x = PM2.5)
) +
  geom_histogram()
```

### R Markdown structure. YAML header and layout

In the YAML header, we should specify the desired title and the output (`flexdashboard::flex_dashboard`) in this case.  Columns are specified using `--------------------`. The components to be included are specified by using `###`. Column width can be specified as 600 pixels using `Column {data-width=600}`. Paste the desired R code below the column name.  


















