---
title: "Notes on Chapter 9"
author: "The Caveman Coder"
date: "2026-01-01"
output: html_document
---

# Spatial modeling of geostatistical data: Malaria in the Gambia

Aim: show how to fit a geostatistical model to predict malaria prevalence in the Gambia using stochastic partial differential equation (SPDE) and plot the results by means of interactive maps created using **leaflet**.  

## Data

Loading the `gambia` dataset from the **geoR** package:
```{r}
library(geoR)
data(gambia)
```

Data dictionary:
- `x`: the x-coordinate of the village (UTM).  
- `y`: the y-coordinate of the village (UTM).  
- `pos`: presence (1) or absence (0) of malaria in a blood sample taken from a child.  
- `age`: age of the child, in days.  
- `netuse`: indicator variable denoting whether the child regularly sleeps under a bed net.  
- `treated`: indicator variable denoting whether the bed net is treated or not.  
- `green`: satellite-derived measure of the greenness of vegetation in the vicinity of the village.  
- `phc`: indicator variable denoting the presence or absence of a health center in the village.  

```{r}
head(gambia)
```

## Data preparation

Data in `gambia` are given at an individual level. For the analysis, this data need to be aggregated by village. 

### Prevalence

Inspecting the data:
```{r}
dim(gambia)
dim(unique(gambia[, c("x", "y")]))
```

Calculating the prevalence of malaria cases:
```{r}
library(dplyr)

d <- gambia |> group_by(x, y) |> 
  summarize(
    total = n(),
    positive = sum(pos),
    prev = positive / total
  )

head(d)
```


### Transforming coordinates

`leaflet()` expects data to be specified in geographic coordinates (long/lat). The coordinates in the `gambia` dataset are in UTM format (easting/northing). 

Transforming the UTM coordinates using `st_transform()` from the **sf** package:
```{r}
library(sf)

sf <- st_as_sf(
  as.data.frame(d), 
  coords = c("x", "y"),
  crs = "+proj=utm +zone=28"
)

sft <- st_transform(sf, crs = 4326)
```

Adding the longitude and latitude variables to dataframe `d`:
```{r}
d[, c("long", "lat")] <- st_coordinates(sft)
head(d)
```


### Mapping prevalence

Creating an interactive map in **leaflet** with circles whose color correspond to the prevalence rates of malaria:
```{r fig.cap="Malaria prevalence in The Gambia."}
library(leaflet)
library(viridis)

pal <- colorBin("viridis", bins = c(0, 0.25, 0.5, 0.75, 1))

leaflet(d) |> 
  addProviderTiles(providers$CartoDB.Positron) |> 
  addCircles(
    lng = ~ long,
    lat = ~ lat,
    color = ~ pal(prev)
  ) |> 
  addLegend(
    "bottomright",
    pal = pal,
    values = ~ prev,
    title = "Prev."
  ) |> 
  addScaleBar(position = c("bottomleft"))
```


### Environmental covariates

Aim: to model malaria prevalence in the Gambia using a covariate that indicates the elevation.

Obtaining the elevation information using the function `elevation_30s()` from the **geodata** package:
```{r}
library(geodata)

r <- elevation_30s(country = "GMB", path = tempdir())
```

The function `elevation_30s()` obtains elevation for any country (specified using the ISO country code) at a resolution of 30 arc seconds (about 1 km).

Adding elevation data to the leaflet plot using `addRasterImage()`:
```{r}
pal <- colorNumeric(
  "viridis",
  values(r),
  na.color = "transparent"
)

leaflet() |> 
  addProviderTiles(providers$CartoDB.Positron) |> 
  addRasterImage(
    r,
    colors = pal,
    opacity = 0.5
  ) |> 
  addLegend(
    "bottomright",
    pal = pal,
    values = values(r),
    title = "Elevation"
  ) |> 
  addScaleBar(position = c("bottomleft"))
```

Adding the elevation values (by extracting them using `terra::extract()`) to dataframe `d` to be able to use it as a covariate in the model:
```{r}
d$cov <- terra::extract(r, d[, c("long", "lat")])$GMB_elv_msk
```

Viewing the final dataset:
```{r}
head(d)
```


## Modeling

Aim: Create a model to predict malaria prevalence in The Gambia using the SPDE approach and the **R-INLA** package.

### Model

Conditional on the true prevalence $P(\mathbf{x}_i)$ at location $\mathbf{x}_i, i = 1, ..., n$, the number of positive results $Y_i$ out of $N_i$ people sampled follows a binomial distribution:
$$
Y_i|P(\mathbf{x}_i) \sim Binomial(N_i, P(\mathbf{x}_i)), \\
logit(P(\mathbf{x}_i)) = \beta_0 + \beta_1 \times \text{elevation} + S(\mathbf{x}_i)
$$

where $\beta_0$ denotes the intercept, and $\beta_1$ denotes the coefficient of elevation,  and $S(\cdot)$ is a spatial random effect that follows a non-zero mean Gaussian process with Matem covariance function...  

### Mesh construction

Similar to what was done in chapter 8, it is necessasry to build a triangulated mesh that covers The Gambia over which to make the random field discretization (is it because the model follows a binomial distribution instead of normal?). 

Building the triangulated mesh using the function `inla.mesh.2d()` from the **R-INLA** package:
```{r}
library(INLA)

coo <- cbind(d$long, d$lat)
# mesh <- inla.mesh.2d(
mesh <- fmesher::fm_mesh_2d_inla(
  loc = coo,
  max.edge = c(0.1, 5),
  cutoff = 0.01
)
```

Note: `inla.mesh.2d()` was deprecated. The newer function is `fmesher::fm_mesh_2d_inla()`.

Inspecting the number of mesh vertices:
```{r}
mesh$n
```

Plotting the mesh object:
```{r fig.cap="Triangulated mesh used to build the SPDE model."}
plot(mesh)
points(coo, col = "red")
```


### Building the SPDE model on the mesh

Using the `inla.spded2.matern()` function to build the SPDE model:
```{r}
spde <- inla.spde2.matern(
  mesh = mesh,
  alpha = 2,
  constr = TRUE
)
```

Here the argument `constr` was set to `TRUE` to impose an integrate-to-zero constraint. `alpha` is a parameter related to the smoothness parameter of the process $\alpha = v + d/2$. If we want the smoothness parameter $v$ to equal 1, the value of `alpha` becomes 2 (d = 2 in our spatial case).

### Index set

Generating the indes set for the SPDE model using `inla.spde.make.index()`:
```{r}
indexs <- inla.spde.make.index("s", spde$n.spde)
lengths(indexs)
```


### Projection matrix
 
Building the projection matrix $A$ that will be used to project the spatially continuous Gaussian random field from the observations to the mesh nodes. 

Using `inla.spde.make.A()` function to create the projection matrix:
```{r}
A <- inla.spde.make.A(mesh = mesh, loc = coo)
```

### Prediction data

Extracting the locations from the raster data (the locations are where the model will predict the prevalence from) using the `crds()` function from **terra**:
```{r}
dp <- terra::crds(r)
dim(dp)
```

There are 12847 points -- these are location points that the model will predict from. 

Lowering the resolution of the raster using the `aggregate()` function from **terra** (to make the computation faster):
```{r}
ra <- terra::aggregate(r, fact = 5, fun = mean)
```

Extracting prediction points from the aggregated `r` raster data:
```{r}
dp <- terra::crds(ra)
dp <- as.data.frame(dp)
dim(dp)
```

Now we have fewer prediction points. 

Extracting the prediction points (location points) again, this time as variable `coop`:
```{r}
coop <- terra::crds(ra)
```

Adding to the prediction data `dp` a column with elevation values for the prediction coordinates:
```{r}
dp$cov <- terra::extract(ra, coop)[, 1]
head(dp)
```

Creating another projection matrix for projevting the spatially continuous Gaussian random field from the prediction locations to the mesh nodes:
```{r}
Ap <- inla.spde.make.A(mesh = mesh, loc = coop)
```

### Stack with data for estimation and prediction

Using the `inla.stack()` function to organize data, effects, and prediction matrices:
```{r}
# stack for estimation
stk.e <- inla.stack(
  tag = "est",
  data = list(y = d$positive, numtrials = d$total),
  A = list(1, A),
  effects = list(data.frame(b0 = 1, elevation = d$cov), s = indexs)
) 

# stack for prediction
stk.p <- inla.stack(
  tag = "pred",
  data = list(y = NA, numtrials = NA),
  A = list(1, Ap),
  effects = list(data.frame(b0 = 1, elevation = dp$cov), s = indexs)
)

# the full stack
stk.full <- inla.stack(stk.e, stk.p)
```

### Model formula

Specifying the model formula by including in the left-hand side the response variable and in the right-hand side the fixed and random effects:
```{r}
formula <- y ~ 0 + b0 + elevation + f(s, model = spde)
```

### `inla()` call

Fitting the model by calling `inla()` and using the default priors in **R-INLA**:
```{r}
res <- inla(
  formula,
  family = "binomial",
  Ntrials = numtrials,
  control.family = list(link = "logit"),
  data = inla.stack.data(stk.full),
  control.predictor = list(
    compute = TRUE,
    link = 1,
    A = inla.stack.A(stk.full)
  ),
  control.compute = list(return.marginals.predictor = TRUE)
)
```

The argument `control.predictor` set to `compute = TRUE` will allow the model to compute the posteriors of the predictors. `link` when set to `1` will allow the model to compute the fitted values (the `res$summary.fitted.values`` and `res$marginals.fitted.values`) with the same link function as the `family` specified.  


## Mapping malaria prevalence

Obtaining the indices that correspond to the prediction locations:
```{r}
index <- inla.stack.index(stack = stk.full, tag = "pred")$data
```

Extracting the vectors with the mean prevalence and the upper and lower limits of 95% credible intervals:
```{r}
prev_mean <- res$summary.fitted.values[index, "mean"]
prev_ll <- res$summary.fitted.values[index, "0.025quant"]
prev_ul <- res$summary.fitted.values[index, "0.975quant"]
```


Creating a map with the predicted prevalence at the prediction locations given in `coop`:
```{r fig.cap="Map of malaria prevalence predictions in The Gambia creatd with `leaflet()` and `addCircles()`."}
pal <- colorNumeric(
  palette = "viridis",
  domain = c(0, 1),
  na.color = "transparent"
)

leaflet() |> 
  addProviderTiles(providers$CartoDB.Positron) |> 
  addCircles(
    lng = coop[, 1], 
    lat = coop[, 2],
    color = pal(prev_mean)
  )  |> 
  addLegend(
    "bottomright",
    pal = pal,
    values = prev_mean,
    title = "Prev."
  ) |> 
  addScaleBar(position = c("bottomleft"))
```

Plotting the prevalence prediction as a raster:
```{r fig.cap="Map of malaria prevalence predictions in The Gambia created with `leaflet()` and `addRasterImage()`."}
# create a SpatVector object
sv <- terra::vect(
  coop,
  atts = data.frame(
    prev_mean = prev_mean,
    prev_ll = prev_ll,
    prev_ul = prev_ul
  ),
  crs = "+proj=longlat +datum=WGS84"
)

# rasterize
r_prev_mean <- terra::rasterize(
  x = sv, y = ra,
  field = "prev_mean",
  fun = mean
)

# plotting the raster with the prevalence values in a map
pal <- colorNumeric("viridis", c(0, 1), na.color = "transparent")

leaflet() |> 
  addProviderTiles(providers$CartoDB.Positron) |> 
  addRasterImage(
    r_prev_mean,
    colors = pal,
    opacity = 0.5
  ) |> 
  addLegend(
    "bottomright",
    pal = pal,
    values = values(r_prev_mean),
    title = "Prev."
  ) |> 
  addScaleBar(position = c("bottomleft"))
```

Using the same approach to create maps with the lower and upper limits of the prevalence predictions:
```{r fig.cap="Map of lower limit of 95% CI of malaria prevalence in The Gambia."}
r_prev_ll <- terra::rasterize(
  x = sv, y = ra,
  field = "prev_ll",
  fun = mean
) 

leaflet() |> 
  addProviderTiles(providers$CartoDB.Positron) |> 
  addRasterImage(
    r_prev_ll,
    colors = pal,
    opacity = 0.5
  ) |> 
  addLegend(
    "bottomright",
    pal = pal,
    values = values(r_prev_ll),
    title = "LL"
  ) |> 
  addScaleBar(position = c("bottomright"))
```

Plot for the upper limit: 
```{r fig.cap="Map of upper limit of 95% CI of malaria prevalence in The Gambia."}
r_prev_ul <- terra::rasterize(
  x = sv, y = ra,
  field = "prev_ul",
  fun = mean
)

leaflet() |> 
  addProviderTiles(providers$CartoDB.Positron) |> 
  addRasterImage(
    r_prev_ul,
    colors = pal,
    opacity = 0.5
  ) |> 
  addLegend(
    "bottomright",
    pal = pal,
    values = values(r_prev_ul),
    title = "UL"
  ) |> 
  addScaleBar(position = c("bottomleft"))
```


## Mapping exeedance probabilities

This answers the question "What are the probabilities that malaria prevalence is greater than a particular threshold?"

Calculating the exeedance probabilities if the threshold is 20%:
```{r}
index <- inla.stack.index(stack = stk.full, tag = "pred")$data

1 - inla.pmarginal(
  q = 0.2,
  marginal = res$marginals.fitted.values[index][[1]]
)
```

Computing the exceedance probabilities for all the prediction locations using the `sapply()` function:
```{r}
excprob <- sapply(
  res$marginals.fitted.values[index],
  FUN = function(marg){1 - inla.pmarginal(q = 0.20, marginal = marg)}
)

head(excprob)
```


Making maps of the exceedance probabilities by creating a raster with exceedance probabilities using `rasterize()` and `leaflet()`:
```{r fig.cap="Map of malaria probability prevalence in The Gambia exceeding 20%"}
sv$excprob <- excprob
r_excprob <- terra::rasterize(
  x = sv, y = ra,
  field = "excprob",
  fun = mean
)

pal <- colorNumeric("viridis", c(0, 1), na.color = "transparent")

leaflet() |> 
  addProviderTiles(providers$CartoDB.Positron) |> 
  addRasterImage(
    r_excprob,
    colors = pal,
    opacity = 0.5
  ) |> 
  addLegend(
    "bottomright",
    pal = pal,
    values = values(r_excprob),
    title = "P(p>0.2)"
  ) |> 
  addScaleBar(position = c("bottomleft"))
```



