---
title: "Notes-Chapter2"
author: "The Caveman Coder"
date: "2025-12-29"
output: html_document
---

# Chapter 2: Spatial data and R packages for mapping

## Types of spatial data

- **areal data:** data domain is partitioned to a finite number of areal units. Examples: attributes collected by ZIP code, or remotely-sensed data reported by pixels.  
- **geostatistical data:** the domain is a continuous fixed set.  
- **point patterns:** data domain is random.    


Example of areal data:
```{r}
library(sf)
library(ggplot2)
library(viridis)

nc <- st_read(
  system.file("shape/nc.shp", package = "sf"),
  quiet = TRUE
)

head(nc)
```

Plot of the example areal data:
```{r}
ggplot(
  data = nc,
  aes(fill = SID74)
) +
  geom_sf() +
  scale_fill_viridis() +
  theme_bw() +
  labs(
    caption = "Sudden infant deaths in North Carolina in 1974."
  )
```

Example of geostatistical data:
```{r}
library(geoR)
library(dplyr)

glimpse(parana)
head(parana, 5)
```

Plot of the example geostatistical data:
```{r}
ggplot(
  data.frame(cbind(parana$coords, Rainfall = parana$data))
) + 
  geom_point(
    aes(east, north, color = Rainfall),
    size = 2
  ) +
  scale_color_gradient(low = "blue", high = "orange") +
  geom_path(
    data = data.frame(parana$border),
    aes(east, north)
  ) +
  theme_bw() +
  labs(
    caption = "Average rainfall measured at 143 recording stations in Parana state, Brazil."
  )
```


Example of point patterns:
```{r}
library(cholera)

head(fatalities)
```

Plot of the example point patterns:
```{r}
rng <- mapRange()

plot(fatalities[, c("x", "y")],
     pch = 15, col = "black",
     cex = 0.5,
     xlim = rng$x,
     ylim = rng$y,
     asp = 1,
     frame.plot = FALSE, 
     axes = FALSE, 
     xlab = "", ylab = ""
)
addRoads()
```
John Snow's map of the 1854 London cholera outbreak. 


## Coordinate reference systems

Locations on Earth can be referenced using unprojected (or geographic) or projected coordinate reference systems.  
- Unprojected/geographic reference systems use longitude and latitude for referencing a location on the Earth's 3d surface.  
- Projected coordinate reference systems use easting northing Cartesian coordinates for referencing a location on a 2d representation of the Earth

### Geographic coordinate systems

- Latitude and longitude are angles given in decimal degrees (DM) or in degrees, minutes, seconds (DMS).  
- The prime meridian passes through the British Royal Observatory in Greenwich, England.  
- The prime meridian determines the eastern and western hemispheres.  
Longitude values range from -180 degrees when running west to 180 degrees when running east.  


### Projected coordinate systems

Since a map projection is a transformation of the Earth's 3d surface as a flat, 2d plane, all map projections distort the Earth's surface in some way.  

The Mercator projection preserves local angles and shapes (useful in navigation especially during the Age of Exploration), but distorts the land area.  

- A variant of the original Mercator projection, called the Universal Transverse Mercator (UTM), divides the Earth into 60 zones of 6 degrees of longitude in width.  
- Each of the zones uses a transverse Mercator projection that maps a region of large north-south extent.  
- A position on Earth is givben by the UTM zone number, the hemisphere, and the easting and northing coordinates in the zone, measured in meters.  


### Setting Coordinate Reference Systems in R

- The Earth's shape is not a perfect sphere. It is an oblate spheroid that bulges at the equator and is flattened at the poles.  
- The most common reference ellipsoid in use is called WGS84 or World Geodetic System.  
- WGS84 is used by the Global Positioning System (GPS).  
- The position of the model ellipsoid relative to the center of the Earth defined by the **datum**. The datum provides the origin point and defines the direction of the coordinate axes.  

Script that defines the WGS84 longitude/latitude projection in R:
```r
"+proj=longlat +ellps=WGS84 +datum=WGS84 +no_defs"
```

Another common reference system is the one used by the European Petroleum Survey Group and is known as the EPSG code.  
- The EPSG equivalent of WGS84 projection is 4326.  
- Other EPSG codes of commonly-used projections can be found on http://www.spatialreference.org.  


## Shapefiles  

A popular storage format for geographic data is the shapefile format, which contains three mandatory files with the following extensions:  
- `.shp`: contains the geometry data,
- `.shx`: contains the positional index of the geometry data.  
- `.dbf`: contains the attributes for each shape.  

Optional files can have the following extension names:  
- `.prj`: plain text file that describes the projection  
- `.sbn` and `.sbx`: contains the spatial index of the geometry data  
- `.shp` and `.xml`: contains geospatial metadata in XML format.  

Reading shapefiles using the function `st_read()` from the **sf** package:
```{r}
# getting the shapefiles for North Carolina
nameshp <- system.file("shape/nc.shp", package = "sf")

# read the shapefile using st_read()
map <- st_read(nameshp, quiet = TRUE)

class(map)
head(map)
```

Plotting different maps from the above sample shapefile (North Carolina):
```{r}
plot(map)
```


## Making maps with R

Mapping with R is commonly done using packages like **ggplot2**, **leaflet**, **mapview**, and **tmap**.

### Mapping with ggplot2

Example:
```{r}
library(sf)
library(ggplot2)

nameshp <- system.file("shape/nc.shp", package = "sf")
map <- st_read(nameshp, quiet = TRUE)
ggplot(map) +
  geom_sf(aes(fill = SID74)) +
  theme_bw() +
  labs(caption = "Map of sudden infant deaths in North Carolina in 1974, created with ggplot2.")
```

The default color scale for discrete variables is `scale_*_hue()`.  

Modifying the plot by using a different color scale (viridis scale):
```{r}
ggplot(map) +
  geom_sf(aes(fill = SID74)) +
  scale_fill_viridis() +
  labs(caption = "Map of sudden infant deaths in North Carolina in 1974, created with ggplot2 and viridis scale.")
```

Saving a plot produced with **ggplot2** is usually done using the `ggsave()` function. 

Alternative way to save the plot: first specify the device driver (`png`, `pdf`, etc.), printing the plot, then shutting down the device using `dev.off()`.
```{r}
png("plot.png")
ggplot(map) + 
  geom_sf(aes(fill = SID74)) +
  scale_fill_viridis() +
  labs(caption = "Map of sudden infant deaths in North Carolina in 1974, created with ggplot2 and viridis scale.") +
  theme_bw()
dev.off()
```

To create animated or interactive plots, we can use the **gganimate** or **plotly** packages, respectively, in combination with **ggplot2**.  


### leaflet

For making interactive maps, **leaflet** is a popular choice. It is an open-source JavaScript-based library. 

### Example using leaflet

Checking the current coordinate reference system using `st_crs()`:
```{r}
st_crs(map)
```

Transforming the CRS of our map to ESPG 4326:
```{r}
map <- st_transform(map, 4326)
st_crs(map)
```

Defining the desired color palette using `colorNumeric()` and plotting the map using `leaflet():
```{r}
library(leaflet)

pal <- colorNumeric("YlOrRd", domain = map$SID74)

leaflet(map) |> 
  addTiles() |> 
  addPolygons(
    color = "white", fillColor = ~ pal(SID74),
    fillOpacity = 1
  ) |> 
  addLegend(pal = pal, values = ~ SID74, opacity = 1) 
```
Map of sudden infant deaths in North Carolina in 1974, created with leaflet.


### mapview

The **mapview** package allows users to very quickly create interactive visualizations (useful during EDA).  

Example of an interactive map created using `mapview()`:
```{r}
library(mapview)
mapview(map, zcol = "SID74")
```

Adding customizations to `mapview()`:
```{r}
library(RColorBrewer)
pal <- colorRampPalette(brewer.pal(9, "YlOrRd"))

mapview(
  map,
  zcol = "SID74",
  map.types = "CartoDB.DarkMatter",
  col.regions = pal
)
```

The function `sync()` can be used to produce a lattice-like view of multiple maps created with mapview or leaflet, that zoom and pan in a synchronized manner.

Example of maps that feature synchronized panning and zooming:
```{r}
library(leafsync)

m74 <- mapview(map, zcol = "SID74")
m79 <- mapview(map, zcol = "SID79")

m <- leafsync::sync(m74, m79)
m
```
Synchronized maps of sudden infant deaths in North Carolina in 1974 and 1979. 


### tmap

The **tmap** package is used to generate thematic maps. To create thematic maps under this package, a base map is first created using the `tm_shape()` function, then layers are added using the `tm_*()` family of functions. Setting the `tmap_mode()` to `"plot"` produces static maps, and setting the former to `"view:` produces interactive maps.

Example using **tmap** to create an interactive map:
```{r}
library(tmap)

tmap_mode("view")
tm_shape(map) +
  tm_polygons("SID74")
```







